Antisense transcripts can quickly evolve to encode proteins

### 2023-03-21 Manuscripts/M2/antisense_genes.tex
new file:   Manuscripts/Figures/M2_main/pORFgain_antisense_scer.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_scerGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_scerWC.svg 	new file:   Manuscripts/Figures/M2_main/pORFloss_antisense_scer.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_scerGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_scerWC.svg 	modified:   Manuscripts/M2/antisense_genes.tex 	modified:   Modeling/antisenseGenes.jl 	modified:   Modeling/antisenseGenes_supplement.jl

\usepackage[usenames,dvipsnames]{xcolor}

@@

We developed mathematical models to estimate the probabilities of ORF emergence and loss, in DNA regions antisense to existing protein coding ORFs. These models are defined by two kinds of probability. The first is the probability of finding a certain kind of DNA sequence, for example an ORF. This stationary probability depends on the nucleotide composition of the DNA region that can be roughly approximated by GC content or by the frequencies of short DNA sequences (k-mers). The second kind of probability describes the mutational change of a sequence to a different kind of sequence. For example, gain or loss of an ORF. This transition probability depends on the mutation rate and mutation bias, in addition to nucleotide composition. We estimate these parameters primarily from the data on the yeast, \textit{Saccharomyces cerevisiae} \citep{scermutrate}. Our choice is motivated by the fact that yeast is a convenient model organism for laboratory experimental studies that can be used to validate theoretical predictions. We also perform analogous analyses using data obtained from \textit{Drosophila melanogaster} \citep[][Supplementary section XX]{drosophilamutrate}.

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{\figdir/pORF_antisense_scer.pdf}
\caption{Antisense ORFs in frame 1 are more likely to exist than intergenic ORFs of identical lengths. Panel \textbf{(A)} shows the probability of antisense ORFs in frame 1 relative to that of intergenic ORFs ($\log_2$ ratio, vertical axis). Line colors indicate the GC-content of the ORFs (yellow = 30\%, blue = 40\%, red = 50\%, purple = 60\%). We do not show antisense ORFs in frames 0 and 2 because their probabilities are identical to that of intergenic ORFs. Panel \textbf{(B)} shows the probability of antisense ORFs relative to that of intergenic ORFs ($\log_2$ ratio, vertical axis), calculated using frequencies of short DNA sequences from yeast genome. Frames 0, 1 and 2 are denoted by dotted, solid and dashed lines, respectively. Horizontal axis in both panels shows the length of the ORFs. For data in both panels, we assume that antisense ORFs overlap completely with the sense ORF.}
\label{pORF}
\end{figure}

Based on these considerations, we determined the probability of finding an antisense ORF. To this end, we calculated the probability of finding a stop codon in the three antisense frames, given the nucleotide composition of the sense ORF, including the condition that no stop codon exists in the sense ORF. Specifically, a stop codon can exist in frame 0 wherever the reverse complementary codons (\texttt{TTA},\texttt{CTA},\texttt{TCA}) exist in the sense ORF. It can exist in frames 1 and 2, overlapping with one of the 128 and 192 dicodons in the sense ORF, respectively. We next calculated the probability of an antisense stop codon based on the probability of the corresponding sense codons and dicodons. We calculated the probability of finding a start codon using the general GC content of the locus, ignoring any effect of the sense ORF. Using the start and stop codon probabilities, we estimated the probability of finding an antisense ORF of different lengths in each of the three frames. We did so for four different values of GC content. We found that frame 1 has the highest likelihood of harboring an antisense ORF in 99.2\% of cases. The only exceptions are ORFs shorter than 39 codons present in a DNA region with a GC content of 60\%. Even for these exceptional cases, the probability of an antisense ORF in frame 1 is no less than 91\% of the corresponding ORF probabilities in the other frames.

To understand whether antisense ORFs are more likely to be found than ORFs in intergenic regions, we compared the probabilities of antisense ORFs with that of intergenic ORFs of identical GC content and lengths. We note that that the probability of intergenic ORFs is not constrained by existing ORFs and is thus purely dependent on nucleotide composition. We found that the probability of antisense ORFs in frames 0 and 2 were identical to that of the corresponding intergenic ORFs, which means that antisense ORFs in frame 1 are more likely to exist than intergenic ORFs except in the 0.8\% cases where the ORFs have a GC content of 60\% and less than 39 codons.

GC content may not accurately describe the nucleotide composition of a locus. Therefore, we also calculated the probability of antisense ORFs using actual codon and dicodon frequencies in yeast ORFs. Likewise, we calculated the probability of intergenic ORFs using the frequencies of DNA trimers in yeast intergenic genome. With this analysis, we found that antisense ORFs longer than 54 and 75 codons in frames 1 and 2, respectively, are more likely to exist than intergenic ORF of the same lengths. For all ORF lengths, antisense ORFs in frame 3 are less likely to exist than intergenic ORFs.

@@

\subsection*{Antisense ORFs are more likely to emerge in, and less likely to be lost from frame 1}

We next analysed how likely it is for antisense ORFs to emerge, when they are not already present. To this end, we calculated gain probability of antisense ORFs in each of the three frames, and under three different intensities of purifying selection. We also calculated the probability of ORF gain in the intergenic regions. We found that antisense ORFs are less likely to emerge in frames 0 and 2, than ORFs in intergenic regions, for all ORF lengths and GC-content. With increasing intensity of purifying selection, the emergence of these antisense ORFs became increasingly less likely than that of intergenic ORFs.

@@

\section*{Data availability}
We performed all calculations using Julia programming language, and all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution. Specifically, our model is implemented in the scripts \texttt{antisenseGenes.jl}, and  \texttt{antisenseGenes\_supplement.jl}, which in turn depend on the script \texttt{nucleotidefuncts.jl} for some basic functions.

>>>>

\usepackage[section]{placeins}
\usepackage[usenames,dvipsnames]{xcolor}

@@

We developed mathematical models to estimate the probabilities of ORF emergence and loss, in DNA regions antisense to existing protein coding ORFs. These models are defined by two kinds of probability. The first is the probability of finding a certain kind of DNA sequence, for example an ORF. This stationary probability depends on the nucleotide composition of the DNA region that can be roughly approximated by GC-content or by the frequencies of short DNA sequences (k-mers). The second kind of probability describes the mutational change of a sequence to a different kind of sequence. For example, gain or loss of an ORF. This transition probability depends on the mutation rate and mutation bias, in addition to nucleotide composition. We estimate these parameters primarily from the data on the yeast, \textit{Saccharomyces cerevisiae} \citep{scermutrate}. Our choice is motivated by the fact that yeast is a convenient model organism for laboratory experimental studies that can be used to validate theoretical predictions. We also perform analogous analyses using data obtained from \textit{Drosophila melanogaster} \citep[][Supplementary section XX]{drosophilamutrate}.

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{\figdir/pORF_antisense_scer.pdf}
\caption{Antisense ORFs in frame 1 are more likely to exist than intergenic ORFs of identical lengths. Panel \textbf{(A)} shows the probability of antisense ORFs in frame 1 relative to that of intergenic ORFs ($\log_2$ ratio, vertical axis). Line colors indicate the GC-content of the ORFs (yellow = 30\%, blue = 40\%, red = 50\%, purple = 60\%). We do not show antisense ORFs in frames 0 and 2 because their probabilities are identical to that of intergenic ORFs. Panel \textbf{(B)} shows the probability of antisense ORFs relative to that of intergenic ORFs ($\log_2$ ratio, vertical axis), calculated using frequencies of short DNA sequences from the yeast genome. Frames 0, 1 and 2 are denoted by dotted, solid and dashed lines, respectively. Horizontal axis in both panels shows the length of the ORFs. For data in both panels, we assume that antisense ORFs overlap completely with the sense ORF.}
\label{pORF}
\end{figure}

Based on these considerations, we determined the probability of finding an antisense ORF. To this end, we calculated the probability of finding a stop codon in the three antisense frames, given the nucleotide composition of the sense ORF, including the condition that no stop codon exists in the sense ORF. Specifically, a stop codon can exist in frame 0 wherever the reverse complementary codons (\texttt{TTA},\texttt{CTA},\texttt{TCA}) exist in the sense ORF. It can exist in frames 1 and 2, overlapping with one of the 128 and 192 dicodons in the sense ORF, respectively. We next calculated the probability of an antisense stop codon based on the probability of the corresponding sense codons and dicodons. We calculated the probability of finding a start codon using the general GC-content of the locus, ignoring any effect of the sense ORF. Using the start and stop codon probabilities, we estimated the probability of finding an antisense ORF of different lengths in each of the three frames. We did so for four different values of GC-content. We found that frame 1 has the highest likelihood of harboring an antisense ORF in 99.2\% of cases. The only exceptions are ORFs shorter than 39 codons present in a DNA region with a GC-content of 60\%. Even for these exceptional cases, the probability of an antisense ORF in frame 1 is no less than 91\% of the corresponding ORF probabilities in the other frames.

To understand whether antisense ORFs are more likely to be found than ORFs in intergenic regions, we compared the probabilities of antisense ORFs with that of intergenic ORFs of identical GC-content and lengths. We note that that the probability of intergenic ORFs is not constrained by existing ORFs and is thus purely dependent on nucleotide composition. We found that the probability of antisense ORFs in frames 0 and 2 were identical to that of the corresponding intergenic ORFs, which means that antisense ORFs in frame 1 are more likely to exist than intergenic ORFs except in the 0.8\% cases where the ORFs have a GC-content of 60\% and less than 39 codons.

GC-content may not accurately describe the nucleotide composition of a locus. Therefore, we also calculated the probability of antisense ORFs using actual codon and dicodon frequencies in yeast ORFs. Likewise, we calculated the probability of intergenic ORFs using the frequencies of DNA trimers in yeast intergenic genome. With this analysis, we found that antisense ORFs longer than 54 and 75 codons in frames 1 and 2, respectively, are more likely to exist than intergenic ORF of the same lengths. For all ORF lengths, antisense ORFs in frame 3 are less likely to exist than intergenic ORFs.

@@

\subsection*{Antisense overlap can facilitate ORF emergence and reduce ORF loss}

\begin{figure}[!b]
\centering
\includegraphics[scale=0.5]{\figdir/pORFgain_antisense_scer.pdf}
\caption{Antisense overlap can facilitate ORF emergence. Panel \textbf{(A)} shows the probability of ORF emergence in the three antisense frames (left to right) relative to that in intergenic regions ($\log_2$ ratio, vertical axis), at different intensities of purifying selection (top to bottom). Line colors indicate the GC-content of the ORFs. Panel \textbf{(B)} shows the ORFs gain probability in the three antisense frames relative to that in intergenic regions ($\log_2$ ratio, vertical axis), calculated using frequencies of short DNA sequences from the yeast genome. Dotted, solid and dashed lines, denote the zero, weak and strong purifying selection, respectively. Horizontal axis in all panels shows the length of the ORFs. For data in both panels, we assume that antisense ORFs overlap completely with the sense ORF.}
\label{pORFgain}
\end{figure}

We next analysed how likely it is for antisense ORFs to emerge, when they are not already present. To this end, we calculated gain probability of antisense ORFs in each of the three frames, and under three different intensities of purifying selection. We also calculated the probability of ORF gain in the intergenic regions. We found that antisense ORFs are less likely to emerge in frames 0 and 2, than ORFs in intergenic regions, for all ORF lengths and GC-content. In contrast, the emergence of antisense ORF emergence in frame 1 is often more likely than that of intergenic ORFs, especially when the ORFs are long.

Increasing the intensity of purifying selection, reduces the emergence likelihood of antisense ORFs in all three frames. However, emergence of antisense ORFs in frame 1 is still more likely than emergence of intergenic ORFs, even under strong purifying selection. Specifically, the minimum ORF length where frame 1 antisense ORFs are more likely to emerge than intergenic ORFs, increases with GC-content and the intensity of selection. For example, in the absence of purifying selection and a GC-content of 30\%, this length is smaller than the smallest investigated ORF length of 30 codons. At the same intensity of selection, this length is 46 codons when the GC-content is 60\%. ORFs can emerge in antisense frame 1 more frequently than in intergenic regions under strong purifying selection, with GC-contents of 30\% and 60\%, if they are longer 49 and 109 codons, respectively.

\begin{figure}[!b]
\centering
\includegraphics[scale=0.5]{\figdir/pORFloss_antisense_scer.pdf}
\caption{Antisense overlap can reduce ORF loss. Panel \textbf{(A)} shows the probability of ORF loss in the three antisense frames (left to right) relative to that in intergenic regions ($\log_2$ ratio, vertical axis), at different intensities of purifying selection (top to bottom). Line colors indicate the GC-content of the ORFs. Panel \textbf{(B)} shows the ORFs loss probability in the three antisense frames relative to that in intergenic regions ($\log_2$ ratio, vertical axis), calculated using frequencies of short DNA sequences from the yeast genome. Dotted, solid and dashed lines, denote the zero, weak and strong purifying selection, respectively. Horizontal axis in all panels shows the length of the ORFs. For data in both panels, we assume that antisense ORFs overlap completely with the sense ORF.}
\label{pORFloss}
\end{figure}

Our analysis of ORF gain probabilities using the frequencies of DNA oligomers (codons, dicodons and intergenic trimers), also shows that antisense ORFs are very likely to emerge in frame 1. ORFs longer than 30 (shortest investigated length), 60 and 69 codons are more likely to emerge in antisense frame 1 than in intergenic regions, when the purifying selection is absent, weak and strong, respectively. Interestingly, this analysis revealed that although antisense ORFs are less likely to emerge in frame 2 than in frame 1, they can emerge more frequently than intergenic ORFs. Specifically when the purifying selection is absent, weak and strong, ORFs that are more likely to emerge in antisense frame 2 than in intergenic regions, contain at least 30, 43 and 82 codons, respectively.

Purifying selection reduces the number of tolerated mutations in a DNA locus. We note again that our lowest level of purifying selection also disallows nonsense mutations from occurring in the sense ORFs. We thus hypothesized that overlap with a sense ORF may protect the antisense ORFs from being lost. To this end, we calculated ORF loss probabilities for different ORF lengths, GC-content, and intensities of purifying selection. In an analogous analysis, we used codon, dicodon, and intergenic trimer frequencies, instead of GC-content, to calculate ORF loss probabilities. Our analyses show that antisense ORFs are indeed protected from loss due to overlap with existing ORFs. That is, the loss probability of antisense ORFs is less than that of intergenic ORFs. This protection against loss increases with increasing intensity of purifying selection. However, selection does not protect the ORFs in all the three antisense frames equally. Antisense ORFs are more likely to be preserved when they exist in frame 1.

Overall, our analyses suggest that antisense overlap with an existing ORF facilitates emergence of new ORFs, and protects the existing antisense ORFs from being lost. A complementary analysis with parameters estimated from \textit{D. melanogaster} further supports this finding.

%frame 1 vs intergenic 
%
%no sel:
% 30
% 30
% 33
% 46
%
%ws: 
% 43
% 54
% 70
% 93
% 
%ss:
%  49
%  63
%  81
% 109

%genomic:
%ns, ws, ss : 30, 60, 69

@@

\section*{Data availability}
We performed all calculations using Julia programming language, and all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution. Specifically, our model uses the following scripts:

\vspace{-2em}
\begin{itemize}\setlength{\itemsep}{-1pt}
\item \texttt{antisenseGenes.jl} (main script)
\item \texttt{antisenseGenes\_supplement.jl} \\(calculations using codon, dicodon, and intergenic trimer frequencies)
\item \texttt{nucleotidefuncts.jl} (dependency for basic functions)
\end{itemize}
